---
title: "Exploratory Data Analysis of Australian Road Fatalities"
author: "Grant Chalmers"
date: "Created 22 Feb, 2020; Last Update `r format(Sys.time(), '%d %B, %Y')`"
output:
  # html_document:
  github_document:
    df_print: paged
    toc: true
    # toc_float: yes
    toc_depth: 3
  
---
## Preparation

### Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  fig.align = "centre",
  fig.height = 6,
  fig.width = 8,
  message = T,
  warning = T,
  error = T)

options(tibble.print_max = 5, tibble.print_min = 5)

# Remove scientific notation
options(scipen=999)

# Set colours for some plots
my_colours <- c("#CC79A7", "#56B4E9", "#009E73", "salmon", "peru", "peachpuff4")

```

```{r load_libraries, include=TRUE, message=FALSE, warning=FALSE,}

# Load required libraries
library(tidyverse)   # core Tidyverse - Loads dplyr & purrr
library(lubridate)   # date & time manipulation
library(ggridges)    # https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html
library(janitor)     # simple functions for examining and cleaning dirty data
library(scales)      # provides internal scaling infrastructure used by ggplot2
library(patchwork)   # makes combining separate ggplots ridiculously simple

# Set a consistent ggplot theme
theme_set(theme_minimal())

```

### Read in data

```{r echo=TRUE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
ardd_fatalities <- read_csv("~/R/ardd/data/ardd_fatalities.csv")
ardd_fatal_crash <- read_csv("~/R/ardd/data/ardd_fatal_crashes.csv")
est_pop <- read_csv("~/R/ardd/data/estimated_population.csv")
state_area <- read_csv("~/R/ardd/data/state_area.csv")

glimpse(ardd_fatalities)
glimpse(ardd_fatal_crash)

# Join datasets
fatal <- ardd_fatal_crash %>% left_join(ardd_fatalities)

```

## Data Description

The Australian Road Deaths Database (ARDD) is maintained and published by the
Bureau of Infrastructure, Transport and Regional Economics (BITRE).
It commenced in 1989 and is updated on a monthly basis.
The ARDD contains basic demographic and crash details of people who have died in
an Australian road crash. Every fatal road traffic crash in Australia is in
scope, and information is included for all people who were killed.

It is published in two forms:

* Fatalities: each record is a killed person
* Crashes: each record is a fatal crash

The database can be found [here](https://bitre.gov.au/statistics/safety/fatal_road_crash_database.aspx),
csv files can be found [here](https://data.gov.au/dataset/ds-dga-5b530fb8-526e-4fbf-b0f6-aa24e84e4277/details?q=crash),
data dictionary can be found [here](https://bitre.gov.au/statistics/safety/files/ARDD_Dictionary_V3.pdf).

Estimated population can be sourced from  [here](https://www.abs.gov.au/AUSSTATS/abs@.nsf/DetailsPage/3101.0Jun%202019?OpenDocument)
(TABLE 4. Estimated Resident Population, States and Territories (Number))

The data is already in a tidy format.

### Data Cleaning

```{r echo=TRUE, data_cleaning}
# Clean if necessary and fix column names via janitor package
fatal <- fatal %>% remove_empty(which = c("rows", "cols")) %>% 
  clean_names()

# Perform overview of variables
glimpse(fatal)

# Create new column to contain hour number, which is easier to analyse, 
# duplicated columns created in join function previously
fatal <- fatal %>% mutate(hour = hour(fatal$time)) %>% 
  select(-ends_with("_2")) %>% glimpse()

# There are quite a few duplicates
# fatal %>% get_dupes(crash_id)
# Remove
fatal_unique <- fatal %>% distinct(crash_id, .keep_all = TRUE)

```

## Visualisations

### Histograms


```{r univariate_plots_1, message=FALSE, warning=FALSE}
# Review total fatality breakdown per crash, exclude extreme outliers on x-axis
hist1 <- fatal_unique %>%
  filter(!is.na(number_fatalities)) %>% 
  ggplot(aes(x = number_fatalities)) +
  geom_histogram(binwidth = 1, fill = my_colours[1], alpha = 0.70) +
  scale_x_continuous(lim = c(0, quantile(fatal_unique$number_fatalities, .999, na.rm = TRUE))) +
  scale_y_continuous() +
  labs(x = "Number of Fatalities", y = "Count",
       title = 'Australian Road Fatalities (1989-2019)') +
  theme(plot.title = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray20"),
        axis.title = element_text(size = 10), legend.position = "bottom",
        axis.text.x = element_text(angle = 0, size = 9),
        strip.text.x = element_text(size = 10, colour = "darkgreen", face = "bold"),
        plot.caption = element_text(size = 8.5, color = "gray50", face = "italic"),
        legend.title = element_text(size = 9), legend.text = element_text(size = 9))
  
# Review count of speed limit zones where a fatality was recorded
hist2 <- fatal_unique %>%
  filter(speed_limit>= 0 & speed_limit <= 130) %>% 
  ggplot(aes(x = speed_limit)) +
  scale_x_continuous(limits = c(10, 130), breaks = seq(10, 130, 10)) +
  geom_histogram(binwidth = 10, fill = my_colours[2], alpha = 0.70) +
  labs(x = "Speed Limit", y = "Count") +
  theme(plot.title = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray20"),
        axis.title = element_text(size = 10), legend.position = "bottom",
        axis.text.x = element_text(angle = 0, size = 9),
        strip.text.x = element_text(size = 10, colour = "darkgreen", face = "bold"),
        plot.caption = element_text(size = 8.5, color = "gray50", face = "italic"),
        legend.title = element_text(size = 9), legend.text = element_text(size = 9))

hist1 / hist2

ggsave("histograms_patchwork.png", plot = last_plot(), path = "images")

```

In the first plot above it is obvious that the vast majority of total fatalities
involved only one fatality. 

From the second plot it is clear that most fatalities occurred in the 60, 80,
100 & 110 km/h speed limit zones, which makes sense considering those are the
most common. 

```{r univariate_plots_2, message=FALSE, warning=FALSE}
# Review count of fatalties by year, excluding 2017
hist3 <- fatal %>% 
ggplot(aes(x = year)) +
  scale_x_continuous(limits = c(1989, 2019), breaks = seq(1989, 2019, 3)) +
  geom_histogram(binwidth = 1, fill = my_colours[1], alpha = 0.70) +
  labs(x = "Year", y = "Count",
       title = 'Australian Road Fatalities (1989-2019)') +
  # scale_fill_manual(values = my_colours) +
  theme(plot.title = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray20"),
        axis.title = element_text(size = 10), legend.position = "bottom",
        axis.text.x = element_text(angle = 45, size = 9),
        strip.text.x = element_text(size = 10, colour = "darkgreen", face = "bold"),
        plot.caption = element_text(size = 8.5, color = "gray50", face = "italic"),
        legend.title = element_text(size = 9), legend.text = element_text(size = 9))

# Review count of fatalties by hour
hist4 <- fatal %>%
  ggplot(aes(x = hour)) +
  scale_x_continuous(limits = c(0, 23), breaks = seq(0, 23, 3)) +
  geom_histogram(binwidth = 1, fill = my_colours[2], alpha = 0.70) +
  labs(x = "Hour", y = "Count") +
  theme(plot.title = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray20"),
        axis.title = element_text(size = 10), legend.position = "bottom",
        axis.text.x = element_text(angle = 0, size = 9),
        strip.text.x = element_text(size = 10, colour = "darkgreen", face = "bold"),
        plot.caption = element_text(size = 8.5, color = "gray50", face = "italic"),
        legend.title = element_text(size = 9), legend.text = element_text(size = 9))

# Order the days correctly
fatal$dayweek <- factor(fatal$dayweek,
                        levels = c("Monday", "Tuesday",
                                   "Wednesday", "Thursday",
                                   "Friday", "Saturday",
                                   "Sunday"))

# Review count of fatalties by Day of the week
bar1 <- ggplot(aes(x = dayweek), data = fatal) +
    geom_bar(fill = my_colours[3], alpha = 0.70) +
  labs(x = "Day of Week", y = "Count") +
  theme(plot.title = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray20"),
        axis.title = element_text(size = 10), legend.position = "bottom",
        axis.text.x = element_text(angle = 00, size = 9),
        strip.text.x = element_text(size = 10, colour = "darkgreen", face = "bold"),
        plot.caption = element_text(size = 8.5, color = "gray50", face = "italic"),
        legend.title = element_text(size = 9), legend.text = element_text(size = 9))

(hist3 | hist4) / bar1

ggsave("histograms2_patchwork.png", plot = last_plot(), path = "images")

```

In the first plot above there is a clear downward trend in Australian road
fatality numbers from 1989 to 2019, which is a very favourable trend.

Note: 2020 was excluded because it is not a full year of data.

The count of fatalities per hour show that most fatalities occur in the
afternoon peak hour. 

The count of fatalities per day of the week shows that considerably more
fatalities occur on weekends, which is not surprising. The increase during the 
week is interesting and maybe a reflection of increased activity, drink driving, or 
general distraction as the weekend approaches.

### Boxplots

```{r echo=TRUE, message=FALSE, warning=FALSE, boxplots}
# Create boxplot to view overall Age and Road_User statistics
box1 <- fatal %>% 
  filter(!is.na(road_user),
         !road_user %in% c("Other/-9"),
         !age == -9) %>% 
  ggplot(aes(x = road_user, y = age)) +
  geom_boxplot(varwidth = F, fill = my_colours[3], alpha = 0.70) +
  coord_flip() +
  scale_y_continuous(breaks = seq(0, 100, 10)) +
  stat_summary(fun.y = mean, geom = 'point', shape = 4) +
  labs(y = "Age", x = "Road User") +
  # scale_fill_manual(values = my_colours) +
  theme(plot.title = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray20"),
        axis.title = element_text(size = 10), legend.position = "bottom",
        axis.text.x = element_text(angle = 0, size = 9),
        strip.text.x = element_text(size = 10, colour = "darkgreen", face = "bold"),
        plot.caption = element_text(size = 8.5, color = "gray50", face = "italic"),
        legend.title = element_text(size = 9), legend.text = element_text(size = 9))

# Focus boxplot on Age and Gender statistics for drivers and riders only
box2 <- fatal %>% filter(!gender == "Unknown",
                 !age == -9, !gender == -9,
                 !is.na(number_fatalities),
                 road_user %in% c("Driver", "Motorcyle rider")) %>%
  ggplot(aes(x = gender, y = age)) +
  geom_boxplot(fill = my_colours[1], alpha = 0.70) +
  coord_cartesian(ylim = c(0, 100)) +
  scale_y_continuous(breaks = seq(0, 100, 10)) +
  stat_summary(fun.y = mean, geom = 'point', shape = 4) +
  labs(x = "Gender", y = "Age",
       title = 'Australian Road Fatalities (1989-2019)') +
  theme(plot.title = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray20"),
        axis.title = element_text(size = 10), legend.position = "bottom",
        axis.text.x = element_text(angle = 0, size = 9),
        strip.text.x = element_text(size = 10, colour = "darkgreen", face = "bold"),
        plot.caption = element_text(size = 8.5, color = "gray50", face = "italic"),
        legend.title = element_text(size = 9), legend.text = element_text(size = 9))


# Boxplot on State and Age to see age differences across states
box3 <- fatal %>% filter(!gender == "Unknown",
                 !age == -9, !gender == -9,
                 !is.na(number_fatalities),
                 road_user %in% c("Driver", "Motorcyle rider")) %>%
  ggplot(aes(x = state, y = age)) +
  geom_boxplot(fill = my_colours[2], alpha = 0.70) +
  coord_cartesian(ylim = c(0, 100)) +
  scale_y_continuous(breaks = seq(0, 100, 10)) +
  stat_summary(fun.y = mean, geom = 'point', shape = 4) +
   labs(x = "State", y = "Age") +
  theme(plot.title = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray20"),
        axis.title = element_text(size = 10), legend.position = "bottom",
        axis.text.x = element_text(angle = 0, size = 9),
        strip.text.x = element_text(size = 10, colour = "darkgreen", face = "bold"),
        plot.caption = element_text(size = 8.5, color = "gray50", face = "italic"),
        legend.title = element_text(size = 9), legend.text = element_text(size = 9))

(box2 + box3) / box1

ggsave("boxplots_patchwork.png", plot = last_plot(), path = "images")
```



```{r road_fatalities_trend}
# Always use fatal_unique when aggregating the number_fatalities field, otherwise
# it will be overstated.
fatal_unique %>%
  filter(year < 2020,
         !is.na(number_fatalities)) %>% 
  group_by(year) %>% 
  summarise(sum_fatal = sum(number_fatalities)) %>% #write_csv("oz_road_fatalities.csv")
  ggplot(aes(x = year, y = sum_fatal)) +
  geom_line(col = "purple", size = 1) +
  geom_smooth(method = 'loess', color = 'green') +
  scale_x_continuous(breaks = seq(1989, 2019, 2)) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Year", y = "Number of Fatalities",
       title = "Australian Road Fatalities (1989-2019)",
       caption = "Source: @GrantChalmers | Australian Road Deaths Database (ARDD)") +
  theme(plot.title = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray20"),
        axis.title = element_text(size = 10), legend.position = "right",
        axis.text.x = element_text(angle = 0, size = 9),
        plot.caption = element_text(size = 9, color = "gray50", face = "italic"),
        legend.title = element_text(size = 9), legend.text = element_text(size = 9))
 
ggsave("oz_road_fatalities.png", plot = last_plot(), path = "images")
```

### Lineplots (trends)


Driver and Motor Cycle Rider Fatalities by Age and Gender.


```{r echo=TRUE, message=FALSE, warning=FALSE, driver_mc_age_gender}

# Overall count of fatalities by Age for Driver and Motor Cycle Rider
fatal %>% filter(!gender == "Unspecified",
                 age >= 16, !gender == -9,
                 road_user %in% c("Driver", "Motorcycle rider")) %>% 
  group_by(age, gender) %>% 
  summarise(f_count = n()) %>%    # must use count, otherwise overstated
  ungroup() %>% 
  arrange(age, gender) %>% 
  ggplot(aes(x = age, y = f_count)) +
  coord_cartesian(xlim = c(16, 100), ylim = c(1, 900)) +
  geom_line(aes(colour = gender), size = 1) +
  scale_x_continuous(breaks = seq(16, 100, 4)) +
  scale_y_continuous(breaks = seq(0, 900, 50)) +
  scale_colour_manual(values = my_colours) +
  labs(x = "Age", y = "Number of Fatalities",
       title = 'Australian Driver and Motor Cycle Rider Fatalities by Age and Gender',
       subtitle = "1989-2019",
       colour = "Gender",
       caption = "Source: @GrantChalmers | Australian Road Deaths Database (ARDD)") +
  theme(plot.title = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray20"),
        axis.title = element_text(size = 10), legend.position = "right",
        axis.text.x = element_text(angle = 0, size = 9),
        plot.caption = element_text(size = 9, color = "gray50", face = "italic"),
        legend.title = element_text(size = 9), legend.text = element_text(size = 9))
 
ggsave("age_gender_fatalities.png", plot = last_plot(), path = "images")
```

Fatalities by road user, age and gender.


```{r echo=TRUE, message=FALSE, warning=FALSE, fatal_road_user_age_gender}

# Overall count of fatalities by road user, age and gender
fatal %>% filter(!gender == "Unspecified",
                 !road_user == "Other/-9",
                 age >= 16, !gender == -9) %>% 
  group_by(age, gender, road_user) %>% 
  summarise(f_count = n()) %>%    # must use count, otherwise overstated
  ungroup() %>% 
  arrange(age, gender, road_user) %>% 
  ggplot(aes(x = age, y = f_count)) +
  geom_line(aes(colour = road_user), size = 1, alpha = 0.75) +
  scale_x_continuous(breaks = seq(16, 100, 4)) +
  scale_y_continuous(breaks = seq(0, 700, 50)) +
  scale_colour_manual(values = my_colours) +
  facet_wrap(~ gender, scales = "free_y") +
  labs(x = "Age", y = "Number of Fatalities",
       title = 'Australian Road Fatalities by Road User, Age and Gender (1989-2019)',
       colour = "Road User",
       caption = "Source: @GrantChalmers | Australian Road Deaths Database (ARDD)") +
  theme(plot.title = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray20"),
        axis.title = element_text(size = 10), legend.position = "bottom",
        axis.text.x = element_text(angle = 90, size = 9),
        strip.text.x = element_text(size = 10, colour = "darkgreen", face = "bold"),
        plot.caption = element_text(size = 8.5, color = "gray50", face = "italic"),
        legend.title = element_text(size = 9), legend.text = element_text(size = 9))
 
ggsave("road_user_age_gender_fatalities.png", plot = last_plot(), path = "images")
```

Australian Road Fatalities by State.

```{r fatalities_by_state_facet, message=FALSE, warning=FALSE}
# Plot line over time with 3 year intervals on the x axis
# because it is incomplete
fatal_unique %>% 
  group_by(year, state) %>% 
  summarise(sum_fatal_state_yr = sum(number_fatalities)) %>%
  ggplot(aes(x = year, y = sum_fatal_state_yr)) +
  geom_line(col = "purple", size = 1) +
  geom_smooth(method = 'loess', color = 'green') +
  scale_x_continuous(breaks = seq(1989, 2019, 3), limits = c(1989, 2019)) + 
  facet_wrap( ~ state, scales = "free_y", ncol = 2) +
  labs(x = "Year", y = "Number of Fatalities",
       title = 'Australian Road Fatalities by State (1989-2019)',
       caption = "Source: @GrantChalmers | Australian Road Deaths Database (ARDD)") +
  theme(plot.title = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray20"),
        axis.title = element_text(size = 10), legend.position = "right",
        axis.text.x = element_text(angle = 45, size = 9),
        strip.text.x = element_text(size = 9, colour = "darkgreen", face = "bold"),
        plot.caption = element_text(size = 9, color = "gray50", face = "italic"),
        legend.title = element_text(size = 9), legend.text = element_text(size = 9))

ggsave("fatalities_by_state.png", plot = last_plot(), path = "images")
```



The plot above shows the number of fatalities per year split by state. Here we
can see a downward trend in every state, although some trends are more
consistent than others. The ACT, NT and TAS seem to be more variable, whereas,
NSW, VIC and SA look more smooth and linear.

Note: the y scale is different for each plot based on population and fatalities.

### Bespoke Visualisations

```{r echo=TRUE, fig.height = 6, fig.width = 10, fatal_rate_by_state}
# Create new dataframe to hold summarised state fatalities
fatal_state <- fatal_unique %>% 
  group_by(state) %>%
  summarise(sum_fatal_state = sum(number_fatalities)) %>%
  ungroup() %>% 
  mutate(state = toupper(state)) %>%    # consistency for joining
  arrange(state)

glimpse(fatal_state)

fatal_state_apop <- fatal_state %>% 
  left_join(est_pop) %>% 
  left_join(state_area) %>% glimpse()

# Create variable to contain the number of years to help calculate yearly
# fatality rate. Remove 1 year as 2017 is incomplete
years <- max(fatal_unique$year) - min(fatal_unique$year) - 1

# Add fatality rate per 100,000 people per year and estimated population density.
fatal_state_apop <- fatal_state_apop %>% 
  mutate(fatal_rate = sum_fatal_state / est_pop * 10 ^ 5 / years,
         est_pop_density = est_pop / area,
         mean_fatal_rate = mean(fatal_rate))

# Review fatality rate per state
plt1 <- fatal_state_apop %>% 
  
  ggplot(aes(x = state, y = fatal_rate)) +
  scale_y_continuous(breaks = seq(0, 25, 5)) +
  geom_bar(stat = 'identity', fill = my_colours[5]) +
  geom_hline(aes(yintercept = mean_fatal_rate), colour="#BB0000",
             linetype="dashed") +
  annotate("text", x = "WA", y = fatal_state_apop$mean_fatal_rate + 0.5,
           label = paste0("Mean == ", round(fatal_state_apop$mean_fatal_rate, 3)),
           parse = TRUE, size = 2.5, colour = "grey40") +
  scale_fill_manual(values = my_colours) +
  labs(x = "State", y = "Fatalities per year for every 100,000 people",
       title = 'Australian Road Fatality Rate by State (1989-2019)') +
  theme(plot.title = element_text(size = 10, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray20"),
        axis.title = element_text(size = 10), legend.position = "right",
        axis.text.x = element_text(angle = 0, size = 9),
        strip.text.x = element_text(size = 9, colour = "darkgreen", face = "bold"),
        plot.caption = element_text(size = 9, color = "gray50", face = "italic"),
        legend.title = element_text(size = 9), legend.text = element_text(size = 9))
  

plt2 <- fatal_state_apop %>% 
  
  ggplot(aes(x = fatal_rate, y = est_pop_density)) +
  scale_y_continuous(trans = log10_trans()) +
  geom_point(aes(colour = fatal_rate), size = 5, alpha = 0.70) + 
  geom_text(aes(label = state), hjust = 0, vjust = 0) +
  geom_line(colour = my_colours[5], linetype = 2) +
  scale_colour_gradient(low = my_colours[3], high = my_colours[4]) +
  labs(x = "Fatalities per year for every 100,000 people",
       y = "Est. Population Density (persons/km2) log scale",
       title = 'Est. Population Density by Fatality Rate',
       caption = "Source: @GrantChalmers | Australian Road Deaths Database (ARDD)") +
  theme(plot.title = element_text(size = 10, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray20"),
        axis.title = element_text(size = 10), legend.position = "none",
        axis.text.x = element_text(angle = 0, size = 9),
        strip.text.x = element_text(size = 9, colour = "darkgreen", face = "bold"),
        plot.caption = element_text(size = 7, color = "gray50", face = "italic"),
        legend.title = element_text(size = 9), legend.text = element_text(size = 9))
  
# Combine plots using patchwork
plt <- plt1 + plt2
plt

# Save file and increase dimensions so all labels fit.
ggsave("fatality_rate.png", path = "images", plot = last_plot(), width = 10, height = 6)

```

Population density is calculated based on ABS estimated population (link provided 
earlier) and land area for each state 
(Population Estimates by Statistical Area Level 2, 2005 to 2015) [here](http://www.abs.gov.au/AUSSTATS/abs@.nsf/DetailsPage/3218.02014-15?OpenDocument).

------

